---
title: "Nonlinear regression for Golf Putting"
author: |
  | Mark Andrews
  | Psychology Department, Nottingham Trent University
  | 
  | \faEnvelopeO\  ```mark.andrews@ntu.ac.uk```
fontsize: 10pt
output:
 beamer_presentation:
  keep_tex: true
  fonttheme: "serif"
  includes:
   in_header: preamble.tex
---

```{r, echo=F}
knitr::opts_chunk$set(echo = F, prompt = F, warning = F, message = F, comment='#>')
# Thanks to 
# https://github.com/ramnathv/slidify/issues/189#issuecomment-15850008
hook1 <- function(x){ gsub("```\n*```r*\n*", "", x) }
hook2 <- function(x){ gsub("```\n+```\n", "", x) }
knitr::knit_hooks$set(document = hook1)
```

```{r}
library(tidyverse)
library(magrittr)
theme_set(theme_classic())
```

# Golf Putting Data

* This data provides the number of putting attempts and the number of successful putts at various distances (in feet) from the hole. 
```{r, eval = F, echo=T}
golf_df <- read_csv('golf_putts.csv')
```
```{r}
golf_df <- read_csv('../data/golf_putts.csv')
```

* The absolute number of attempts and successes at each distance is vital information and so ideally we should base our analysis on this data, using a binomial logistic regression or a related model. 
* However, for simplicity here, we will just use the relative frequencies of successes at each distance.
```{r, echo=T}
golf_df %<>% mutate(prob = success/attempts)
```

# Golf Putting Data

```{r}
ggplot(golf_df,
       mapping = aes(x = distance, y = prob)
) + geom_point() + 
  ylab('Probability of the golf ball landing in hole') +
  xlab('Distance from the hole (ft)')
```

# A model
\begin{figure}
\begin{center}
\includegraphics[width=0.9\textwidth]{golf_putting_v2.pdf}
\caption{A golf ball of radius $r$ (left) and the golf hole of radius $R$ (right). The centres of these two circles are $d$ apart. If the golf ball travels in a straight vertical line to the hole, it will fall in. If its trajectory deviates, either to the right or to the line, greater than an angle of $\theta_{\text{crit}}$, it will miss.
The angle $\theta_{\text{crit}}$ is the angle between the vertical line of length $d$ and the tangent line from the centre of the ball to the hole. The line from the centre of the hole meets the tangent line at a right angle. As such, $\theta_{\text{crit}} = \sin^{-1}\left(\tfrac{R}{d}\right)$.
}
\label{golf_diagram}
\end{center}
\end{figure}

# A physical model

The probability that the angle of their putt will be between $0$ and $\theta_{\text{crit}}$ is
$$
\Prob{0 < \theta \leq \theta_{\text{crit}}} = \Phi(\theta_{\text{crit}}\given 0, \sigma^2) - \tfrac{1}{2},
$$
where
$$
\Phi(\theta_{\text{crit}}\given 0, \sigma^2) \triangleq
\int_{-\infty}^{\theta_{\text{crit}}} N(\theta \given 0, \sigma^2),
$$
which is the value at $\theta_{\text{crit}}$ of the cumulative distribution function of a normal distribution of mean 0 and standard deviation $\sigma$.
We simply double the quantity $\Phi(\theta_{\text{crit}}\given 0, \sigma^2) - \tfrac{1}{2}$ to get $\Prob{\theta_{\text{crit}} < \theta \leq \theta_{\text{crit}}}$.

Therefore, the probability of a sucessful putt is
$$
2 \Phi\left(\sin^{-1}\left(\tfrac{R}{d}\right)\given 0, \sigma^2\right) - 1.
$$

# Implementation 

* This is a nonlinear parameteric function of distance $d$, where $R$ is known to have a value of 53.975mm, and $\sigma$ is the single unknown parameter.

* This nonlinear function is easily implemented as follows.
```{r, echo=T}
successful_putt_f <- function(d, sigma){
  R <- 0.17708333 # 53.975mm in feet
  2 * pnorm(asin(R/d), mean=0, sd=abs(sigma)) -1
}
```

# Using `nls`

The `nls` based model using this `successful_putt_f` function is as follows.
```{r, echo=T}
M_putt <- nls(prob ~ successful_putt_f(distance, sigma),
              data = golf_df,
              start = list(sigma = 0.1)
)
```

# The predictions

```{r}
golf_df %>%
  mutate(prediction = predict(M_putt)) %>%
ggplot(mapping = aes(x = distance)) +
  geom_point(mapping = aes(y = prob)) +
  geom_line(mapping = aes(y = prediction), col='red') +
  ylab('Probability of the golf ball landing in hole') +
  xlab('Distance from the hole (ft)')
```

# An simpler alternative?

```{r, echo=T}
M_putt_exp <- nls(prob ~ a + b * exp(-beta * distance),
                  data = golf_df,
                  start = list(a = 0, b = 1, beta = 1)
)
```

